New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-10-14 jso.qmd
go though the paper before submission

---
title: New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{assumption}{\protect\assumptionname}
   - \newtheorem{claim}{\protect\claimname}
   - \newtheorem{condition}{\protect\conditionname}
   - \newtheorem{conjecture}{\protect\conjecturename}
   - \newtheorem{cor}{\protect\corollaryname}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{example}{\protect\examplename}
   - \newtheorem{lem}{\protect\lemmaname}
   - \newtheorem{notation}{\protect\notationname}
   - \newtheorem{problem}{\protect\problemname}
   - \newtheorem{prop}{\protect\propositionname}
   - \newtheorem{rem}{\protect\remarkname}
   - \newtheorem{thm}{\protect\theoremname}
   - \providecommand{\assumptionname}{Assumption}
   - \providecommand{\claimname}{Claim}
   - \providecommand{\conditionname}{Condition}
   - \providecommand{\conjecturename}{Conjecture}
   - \providecommand{\corollaryname}{Corollary}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\examplename}{Example}
   - \providecommand{\lemmaname}{Lemma}
   - \providecommand{\notationname}{Notation}
   - \providecommand{\problemname}{Problem}
   - \providecommand{\propositionname}{Proposition}
   - \providecommand{\remarkname}{Remark}
   - \providecommand{\theoremname}{Theorem}
nocite: |
   @DLMF
---

@@

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index functions
in the literature. @sec-PP-properties introduces the metrics that
measure different properties of PP indexes, smoothness and
squintability, and @sec-JSO describes the new JSO to be used for the
PPGT. @sec-sim-deets summarises the results of two simulation
experiments done to assess JSO's performance: one comparing JSO's
performance improvements relative to an existing optimiser, Creeping
Random Search (CRS), and the other studying the impact of different PP
index properties on optimisation performance. @sec-sim-res presents the
results. @sec-discussion discusses implementation details and insights
for practitioners. @sec-conclusion summarises the work and provides
suggestions for future directions.

@@

\begin{defn}\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}

@@

\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)\ ,\label{eq:matern}
\end{equation}
where $\left\Vert u\right\Vert$ is the Euclidean norm of $u\in\mathbb{R}^{p{\times}d}$, $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function [DLMF 10.25].
\end{defn}

@@

Here the formal definitions of projection distance and squint angle are
given, before the definition of squintability. Two approaches to compute
this metric numerically are described.

\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbb{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $r(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squint angle]\label{def:squint-angle}
Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.

@@

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r(A, A^*)) = f(XA)$. For brevity, denote $g(r)$ as $g(r(A, A^*))$. The squintability of an index function $f(XA)$ is defined as

@@

<!-- \textcolor{orange}{See whether further details are needed for the nonparametric approach.} -->

<!--To the best of our knowledge, -->

<!-- ## Speed -->

<!-- The speed of optimizing an index function can be measured empirically by -->

<!-- recording the duration of the optimisation. Alternatively, one can also -->

<!-- gauge the speed in terms of computational complexity of evaluating the -->

<!-- index function (for instance, in big O notation, with respect to the -->

<!-- sample size) of computing the index function. -->

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space, aiming to balance exploration
(searching new areas) and exploitation (focusing on promising areas).
The algorithm aims to find the optimal solution by adapting the
behaviour of jellyfish to navigate towards the best solution over
iterations [@chou_novel_2021].

@@

# Assessing optimisers: a simulation study {#sec-sim-deets}

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) [@RJ-2021-105; @laa_using_2020] used in the PP
guided tour to explore JSO's behaviour under different hyper-parameter
and data dimension combinations. The second simulation studies the
effect of index properties (smoothness and squintability), along with
JSO hyper-parameters, and data dimension, on the success rate of the JSO
performance. This section describes the simulation details, with the
results deferred to @sec-sim-res.

@@

The performance of JSO is investigated both in comparison to the
existing optimizer, CRS, and across various hyper-parameter values. The
performance is measured by the success rate, defined as the proportion
of simulations that achieves a final index value within 0.05 of the best
index value found among all 50 simulations (see @fig-success-rate for an
illustration). This comparison is based on projection pursuit to find
the pipe shape investigated by @laa_using_2020 using the `holes` index.

Fifty simulations are conducted with both JSO and CRS, in four different
data dimensions ($d = 6, 8, 10, 12$). JSO uses 100 jellyfish with a
maximum of 100 tries, while the CRS allows a maximum of 1000 samples at
each iteration before the algorithm terminates. The different numbers
account for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in @RJ-2021-105 for assessing JSO, where the design
parameters are stored along with index value, projection basis, random
seed, and computation time.

Fifty additional simulations are conducted for each hyper-parameter
combination to analyze how they affect the JSO success rate. This
includes variations in the number of jellyfish (20, 50, and 100) and the
maximum number of tries (50 and 100).

@@

## Factors affecting JSO success rate: index properties and jellyfish hyper-parameters {#sec-app-2}

To assess JSO's performance across various scenarios, two different data
shapes, pipe and a sine wave, are investigated in 6D and 8D spaces using
six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`,
and `stringy`, with varied JSO hyper-parameters. A total of
`r nrow(sim_df)` combinations result, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. Again,
JSO is run 50 times to calculate the success rate for each projection
pursuit.

@@

To construct a relationship among success rate, index properties
(smoothness and squintability), and jellyfish hyper-parameters, a
generalised linear model is fitted using a binomial family and a logit
link function. The data is pre-processed by 1) scaling the JSO
hyper-parameters by a factor of 10 for interpretation, 2) creating a new
binary variable `long_time` to indicate cases with an average run time
over 20 seconds, and 3) re-coding the success rate for the `stringy`
index as 0, because none of the 50 simulations correctly identified the
sine-wave shape.

```{r}
#| label: fig-smoothness
#| out.width: 100%
#| fig.cap: "Illustration of steps to calculate smoothness. For a given projection pursuit problem defined by the shape to find, data dimension and the index function, 1) sample random bases given the orthonormality contraint and calculate the index value associated with each random basis, and 2) fit a Gaussian process model of index values against projection distances to obtain the smoothness measure. "
# basis_smoothness <- sample_bases(idx = "splines2d", n_basis = 1000)
# basis_smoothness |>
#    ggplot(aes(x = dist, y = index)) +
#    geom_point() + 
#    labs(x = "Projection distance", y = "Index value") + 
#    theme_bw() +   
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "smoothness.png"))
```

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: |
#|   Illustration of steps to calculate squintability. For a given projection pursuit problem defined by the shape to find, data dimension and the index function, 1) sample random bases given the orthonormality and projection distance contraint, 2) interpolate the sampled bases to the optimal basis and calculate the projection distance and the index value for each interpolated basis. 3) bin the index values by projection distances to obtain the average index value for each bin, 4) fit the scaled sigmoid function in equation \eqref{eq-squintability} to the binned index values against projection distances using non-linear least square, 5) calculate the squintability measure using equation (@eq-squintability-parametric) with parameters estimated from the model.
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

@@

Performance is assessed based on the final projections across two
optimisers (JSO and CRS) and the success rate across JSO
hyper-parameters. The final projections found by the two optimisers (JSO
and CRS) are presented in @fig-proj, broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50% of cases. Compared to CRS, JSO achieves higher index values and
clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.

## Factors affecting JSO success rate: index properties and jellyfish hyper-parameters

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. The success rate calculated at each hyper-parameter
combination (number of jellyfish and the maximum number of tries) is
presented in @fig-proportion. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For simpler
problems (6 dimensions), small parameter values (20 jellyfish and a
maximum of 50 tries) can already achieve a high success rate. However,
larger parameter values (i.e. 100 jellyfish and a maximum of 100 tries)
are necessary for higher-dimensional problems (8, 10, and 12
dimensions). Increasing both parameters enhances the performance of JSO,
but it also extends the computational time required for the
optimisation, which can be computationally intensive when evaluating the
index function (such as scagnostic indexes) multiple times across
numerous iterations.

@@

The index properties, smoothness and squintability, numerically
characterise the complexity of projection pursuit index, and hence
potential optimisation problems. @tbl-smoothness-squintability presents
the parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column $\nu$ is used as the smoothness metric and the
column $\varsigma$ is calculated as equation
(@eq-squintability-parametric) as the squintability metric.

@@

The simulation study demonstrated that these metrics can be used to
assess a PP index, and compare optimisers. Here we explain how they can
be used for new index and optimiser development. Using the JSO optimiser
for PPGT requires a change in user behaviour. For all the existing
optimisers a single optimisation path is followed. The JSO has multiple
optimisation paths, and needs additional tools to incorporate into the
PPGT, as explained here.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour. Since multiple tour
paths are generated by the jellyfish, the animation function
`animate_*()` will not render the tour path for JSO. To visualise path
visited by each individual jellyfish, assign the animation to an object
(`res <- animate_xy(...)`). This will save the bases visited by the
optimiser, along with relevant metadata, as a tibble data object (see
@RJ-2021-105 for more details on the data object). The bases visited for
individual jellyfish can then be extracted and viewed using the
`planned_tour()` function.

## Computing the metrics for your new index

The `ferrn` package [@RJ-2021-105] provides functionality for computing
the smoothness and squintability metrics. Both cases require sampling
random bases using the `sample_bases()` function, followed by
calculating the specific metric with either `calc_smoothness()` or
`calc_squintability()`.

@@

To sample bases for calculating smoothness, four arguments are required:
the index function, the dataset, the number of random bases to sample,
and the best basis to compute projection distance. The output of
`sample_bases()` is a data frame containing columns/ list-columns of
sampled basis matrix, calculated projection distance and index value.
Parallellasation is available through the `parallel = TRUE` argument to
speed up the computation.

The `calc_smoothness()` function takes the output from `sample_bases()`
and fits a Gaussian process model to the index values against the
projection distances to compute the smoothness metric. Starting
parameters and additional Gaussian process arguments can be specified
and details of the fit can be accessed through the `fit_res` attribute
of the `calc_smoothness()` output.

@@

Sampling bases for calculating squintability includes an additional step
of interpolating between the sampled bases and the best basis. This is
specified through setting the `step_size` and `min_proj_dist` arguments
in the `sample_bases()` function to non-NA numerical values. Since the
projection distance typically ranges from 0 to 2, it is recommended to
set `step_size` to 0.1 or lower, and `min_proj_dist` to be at least 0.5
to ensure a meaningful interpolation length.

The `calc_squintability()` function computes squintability using two
methods: 1) parametrically, by fitting a scaled sigmoid function through
non-linear least square (`method = "nls"`), and 2) non-parametrically,
using kernel smoothing (`method = "ks"`). A `bin_width` argument is
required to average the index values over the projection distance before
the fit. The output provides the estimated parameters for the scaled
sigmoid function ($\theta_1$ to $\theta_4$) and the squintability metric
for the parametric case. For the non-parametric case, it shows the
maximum gradient attained (`max_d`), the corresponding projection
distance (`max_dist`), and the squintability metric as their products.

@@

>>>>

---
title: New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Austin
          country: United States
          postal-code: 78751
    attributes:
        corresponding: true
  - name: Dianne Cook
    email: dicook@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
  - name: Nicolas  Langrené
    email: nicolaslangrene@uic.edu.cn
    affiliations:
        - id: 3
          name: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science, BNU-HKBU United International College
          department: Department of Mathematical Sciences
          city: Zhuhai
          country: China
          postal-code: 519087
  - name: Jessica Wai Yin Leung
    email: Jessica.Leung@monash.edu
    affiliations:
        - id: 2
          name: Monash University
          department: Department of Econometrics and Business Statistics
          city: Melbourne
          country: Australia
          postal-code: 3800
abstract: |
  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 
keywords: 
  - projection pursuit
  - jellyfish search optimiser (JSO)
  - optimisation
  - grand tour
  - high-dimensional data
  - exploratory data analysis
date: last-modified
bibliography: bibliography.bib
format:
  tandf-pdf:
    keep-tex: true
    fontsize: 12pt
    linestretch: 2
editor: 
  markdown: 
    wrap: 72
crossref: 
  eq-prefix: ""
editor_options: 
  chunk_output_type: console
header-includes:
   - \usepackage{algorithm}
   - \usepackage{float} 
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{amssymb}
   - \theoremstyle{plain}
   - \newtheorem{defn}{\protect\definitionname}
   - \newtheorem{prop}{\protect\propositionname}
   - \providecommand{\definitionname}{Definition}
   - \providecommand{\propositionname}{Proposition}
nocite: |
   @DLMF
---

@@

The paper is structured as follows. @sec-background introduces the
background of the PPGT, reviews existing optimisers and index functions
in the literature. @sec-PP-properties introduces the metrics that
measure two properties of PP indexes, smoothness and squintability, and
@sec-JSO describes the new JSO to be used for the PPGT. @sec-sim-deets
outlines two simulation experiments to assess JSO's performance: one
comparing JSO's performance improvements relative to an existing
optimiser, Creeping Random Search (CRS), and the other studying the
impact of PP index properties on optimisation performance, and
@sec-sim-res presents the results. @sec-discussion discusses the
implementation of JSO in the `tourr` package and the PP property
calculation in the `ferrn` package. @sec-conclusion summarises the work
and provides suggestions for future directions.

@@

\begin{defn}[sobolev space]\label{def:sobolev_space}
The Sobolev space $W^{k,p}(\mathbb{R})$ for $1\leq p\leq \infty$ is the set of all functions $f$ in $L^p(\mathbb{R})$ for which all weak derivatives $f^{(\ell)}$ of order $\ell\leq k$ exist and have a finite $L^p$ norm.
\end{defn}

@@

\begin{defn}[Matérn covariance function]\label{def:matern}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)\ ,\label{eq:matern}
\end{equation}
where $\left\Vert u\right\Vert$ is the Euclidean norm of $u\in\mathbb{R}^{p{\times}d}$, $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function [DLMF 10.25].
\end{defn}

@@

This section defines the projection distance and introduces the
squintability metric, followed by a description of two numerical
approaches for computing squintability.

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squint angle]\label{def:squint-angle}

Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.
\end{defn}

@@

\begin{defn}[squintability]\label{def:squintability}
Let $g: \mathbb{R} \mapsto  \mathbb{R}$ be a decreasing function that maps the projection distance $r(A, A^*)$ to the index value $f(XA)$, such that $g(r) = g(r(A, A^*)) = f(XA)$.  The squintability of an index function $f$ is defined by

@@

@@

The Jellyfish Search Optimiser (JSO) mimics the natural movements of
jellyfish, which include passive and active motions driven by ocean
currents and their swimming patterns, respectively. In the context of
optimization, these movements are abstracted to explore the search
space, aiming to balance exploration (searching new areas) and
exploitation (focusing on promising areas). The algorithm aims to find
the optimal solution by adapting the behaviour of jellyfish to navigate
towards the best solution over iterations [@chou_novel_2021].

@@

# Assessing optimisers: two simulation studies {#sec-sim-deets}

This section describes the details of two simulation studies: one
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) [@RJ-2021-105; @laa_using_2020] to explore JSO's behaviour
under different data dimensions. The second simulation first examines
the JSO performance under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries), and then studies
the effect of index properties (smoothness and squintability), along
with JSO hyper-parameters, and data dimension, on the JSO performance
through a logistic regression.

The performance of JSO is measured by the success rate, defined as the
proportion of simulations that achieves a final index value within 0.05
of the best index value found among all 50 simulations (see
@fig-success-rate for an illustration). This comparison is based on
projection pursuit to find the pipe shape investigated by
@laa_using_2020 using the `holes` index.

@@

The JSO performance is investigated both in comparison to the existing
optimizer, CRS, in a pipe-finding problem using the holes index in the
6, 8, 10, and 12-dimensional space. Fifty simulations are conducted in
four data dimensions ($d = 6, 8, 10, 12$) with each optimiser. JSO uses
100 jellyfish with a maximum of 100 tries, while the CRS allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers account for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in @RJ-2021-105 for
assessing JSO, where the design parameters are stored along with index
value, projection basis, random seed, and computation time.

@@

## Factors affecting JSO success rate: JSO hyper-parameters and index properties  {#sec-app-2}

To examine JSO's performance across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. The success rate is
calculated for each combination.

To assess JSO's performance across various PP problems, two different
data shapes, pipe and a sine wave, are investigated in 6D and 8D spaces
using six different PP indexes: `dcor2d_2`, `loess2d`, `MIC`, `TIC`,
`spline`, and `stringy`, under varied JSO hyper-parameters. This results
in a total of `r nrow(sim_df)` scenarios, comprising of
`r nrow(filter(sim_df, index == "holes"))` computed on the pipe data and
`r nrow(filter(sim_df, index != "holes"))` on the sine-wave data. Again,
JSO is run 50 times for each scenario to calculate the success rate for
each PP problem.

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.

```{r}
#| label: fig-smoothness
#| out.width: 100%
#| fig.cap: "Steps for calculating smoothness in a projection pursuit problem. Given the target shape, data dimension and the index function: (1) sample random bases given the orthonormality contraint and calculate their corresponding index values, and (2) fit a Gaussian process model of index values against the sampled bases to obtain the smoothness measure."
# basis_smoothness <- sample_bases(idx = "splines2d", n_basis = 1000)
# basis_smoothness |>
#    ggplot(aes(x = dist, y = index)) +
#    geom_point() + 
#    labs(x = "Projection distance", y = "Index value") + 
#    theme_bw() +   
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "smoothness.png"))
```

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: |
#|   Steps for calculating squintability in a projection pursuit problem. Given the target shape, data dimension and the index function: (1) sample random bases given the orthonormality and projection distance contraint, (2) interpolate the sampled bases to the optimal basis to calculate the projection distance and the index value. (3) bin the index values by projection distances to obtain the average index value for each bin, (4) fit the scaled sigmoid function in equation \eqref{eq-squintability} to the binned index values against projection distances using non-linear least square, 5) calculate the squintability measure using equation (@eq-squintability-parametric) with parameters estimated from the model.
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

@@

Performance of JSO is assessed based on the final projections across the
two optimisers: JSO and CRS. The final projections found by the two
optimisers are presented in @fig-proj, broken down by 10th quantile,
faceted by the data dimensions. In the 6-dimensional data scenario, JSO
consistently identifies a clear pipe shape. The CRS also finds the pipe
shape but with a wide rim, suggesting a further polish search may be
required. With increasing dimensions, JSO may not always identify the
pipe shape due to random sampling, but it still finds the pipe shape in
over 50% of cases. Compared to CRS, JSO achieves higher index values and
clearer pipe shapes across all quantiles in data of 8, 10, 12
dimensions, suggesting its advantage in exploring high-dimensional
spaces.

## Factors affecting JSO success rate: JSO hyper-parameters and index properties

The effect of JSO hyper–parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in @fig-proportion.
The uncertainty is quantified through 500 bootstrap replicates for each
case. As the number of jellyfish and maximum tries increase, the success
rate also increases. For simpler problems (6 dimensions), small
parameter values (20 jellyfish and a maximum of 50 tries) can already
achieve a high success rate. However, larger parameter values (i.e. 100
jellyfish and a maximum of 100 tries) are necessary for
higher-dimensional problems (8, 10, and 12 dimensions). Increasing both
parameters enhances the performance of JSO, but it also extends the
computational time required for the optimisation, which can be
computationally intensive when evaluating the index function (such as
scagnostic indexes) multiple times across numerous iterations.

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. @tbl-smoothness-squintability presents the parameters
estimated from the Gaussian process (outputscale $\eta$, lengthscale
$\ell$, smoothness $\nu$, and nugget $\sigma$) and from the scaled
logistic function ($\theta_1$ to $\theta_4$) for calculating smoothness
and squintability in each PP problem considered. The column $\nu$ is
used as the smoothness metric and the column $\varsigma$ is calculated
as equation (@eq-squintability-parametric) as the squintability metric.

@@

@@

The simulation studies have compared the JSO with CRS and shown how
smoothness and squintability affect the JSO success rate. Here we
explain how they can be used for new index and optimiser development.
Using the JSO optimiser for PPGT requires a change in user behaviour.
For all the existing optimisers a single optimisation path is followed.
The JSO has multiple optimisation paths, and needs additional tools to
incorporate into the PPGT, as explained here.

@@

To use JSO for PPGT in the `tourr` package, specify
`search_f = search_jellyfish` in the guided tour. Unlike existing
optimisers, the animation function `animate_*()` won't directly render
the tour path for JSO due to the generation of multiple tour paths. To
visualise the path visited by each individual jellyfish, assign the
animation to an object (`res <- animate_xy(...)`). This will save the
bases visited by JSO, along with relevant metadata, as a tibble data
object (see @RJ-2021-105 for more details on the data object). The bases
visited for individual jellyfish can then be extracted and viewed using
the `planned_tour()` function.

## Computing the index properties for your new index

The `ferrn` package [@RJ-2021-105] provides functionality for computing
the smoothness and squintability metrics. Both metrics require sampling
random bases using the `sample_bases()` function, followed by the metric
calculation with `calc_smoothness()` or `calc_squintability()`.

@@

To sample bases for calculating smoothness, the following arguments are
required: the index function, the dataset, and the number of random
bases to sample. The output of `sample_bases()` is a data frame with a
list-column of sampled basis matrix and index value. Parallellasation is
available to speed up the index value computation through the
`parallel = TRUE` argument.

The `calc_smoothness()` function takes the output from `sample_bases()`
and fits a Gaussian process model to the index values against the
sampled basis, as the location, to obtain the smoothness metric.
Starting parameters and additional Gaussian process arguments can be
specified and details of the fit can be accessed through the `fit_res`
attribute of the output.

@@

Bases sampling for squintability includes an additional step of
interpolating between the sampled bases and the best basis. This step is
performed when the arguments `step_size` and `min_proj_dist` in the
`sample_bases()` function are set to non-NA numerical values. Given the
projection distance typically ranging from 0 to 2, it is recommended to
set `step_size` to 0.1 or lower, and `min_proj_dist` to be at least 0.5
to ensure a meaningful interpolation length.

The `calc_squintability()` function computes squintability using two
methods: 1) parametrically, by fitting a scaled sigmoid function through
non-linear least square (`method = "nls"`), and 2) non-parametrically,
using kernel smoothing (`method = "ks"`). A `bin_width` argument is
required to average the index values over the projection distance before
the fit. For the parametric case, the output provides the estimated
parameters for the scaled sigmoid function ($\theta_1$ to $\theta_4$)
and the calculated squintability metric as Equation
@eq-squintability-parametric. For the non-parametric case, it shows the
maximum gradient attained (`max_d`), the corresponding projection
distance (`max_dist`), and the squintability metric as their products.

@@

One future direction of this work is to test the JSO, along with its
variations and other swarm-based optimisers, on a broader range of
indexes, for example, scagnostic indexes.
